---
title: "R Basics"
output: 
html_notebook: default
---

# Introduction

Welcome to your first notebook file! Whenever you create a notebook, you'll see some notation appear at the top, including a "title" in double quotes. This one says "R Basics". You can change these settings or not. All notebook files can output to an html file (like the Introduction-to-R.html file you just opened; that was created using a notebook) and so have these notations.

To start, there are a bunch of packages you need to install (at the conference, these have already been installed for you). If you're working through this from home, run the following code chunk (it only needs to be run once for your computer):
```{r}
install.packages(c("tidyverse", "lubridate", "janitor", "readxl", "RSocrata"))
```


Now that you've gone over the `Introduction-to-R` file, we'll start with some of the basics of the R language.

Below you'll practice writing some basic R. When you write a line of code, you can run it by putting your cursor on that line and typing `cmd+return`, which will run one string of code (note: sometimes your code will cover more than one line. This will run a *complete* set of code that has a natural stopping point. This will likely make more sense later). 
You can also run code in a code chunk by using the green "play" button in the upper-right corner. This will run *all* the code in a chunk.

Create a *character vector* that lists your five favorite foods (or movies, TV shows, books, albums, whatever) and then print that vector. 
Remember that you have to give it a name.
```{r}

```


Create a *numeric vector* with every whole number between 10 and 20. 
```{r}

```


Sum up all the numbers in the vector you just created. 
```{r}

```


Create your first code chunk using `cmd+option+i`. In it, create two variables that are each numbers, then add them together.


